Automatic Bicycle Balance Assistance Reduces Probability of Falling at Low Speeds When Subjected to Handlebar Perturbations\\Version 4.dev

### 2025-11-24 main.tex
Jason's final edit pass.

\section*{Highlights}
%
\begin{itemize}
  \item A bicycle steering motor with automatic roll rate feedback control can
    stabilize the bicycle-rider system at speeds as low as 8~\si{\kph}.
  \item Every bicycle rider has their own perturbation resistance threshold
    when subjected to mechanical handlebar perturbations.
  \item Automatic balance assistance reduces the probability of falling while
    riding at 6~\si{\kph} when mechanically perturbed.
  \item Balance assist technology can reduce natural single-actor falls around
    the limits of the rider's control authority.
\end{itemize}

@@

Many more automatically balanced single-track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in human balancing and what effect it may have towards
single-actor falls. Most of these robotic bicycle and motorcycle designers did
not intend for a human rider to also control the stabilized vehicle.
Nevertheless, an automatically stabilized bicycle can be controlled by a human
rider if the motor controlled steer torque and the rider applied steer torque
act in parallel. The effect of this automatic control gives the ability to
effectively change the human-controlled plant dynamics, up to some limits. In
our prior study~\cite{Alizadehsaravi2023}, we demonstrate reduced motion during
distractions and light perturbations due to the balance assist system but
Hanakam, Wehner, and Wrede~\cite{Hanakam2023} recently showed rider
dissatisfaction with the stabilization of a similar vehicle so the overall
possible benefits are not yet definitively established.

The design of our balance assist system relies on the linear Carvallo-Whipple
bicycle model~\cite{Carvallo1899,Whipple1899} which is the simplest bicycle
model that exhibits both the non-minimum phase behavior ``countersteering'' and
self-stability. The bicycle model is well suited for showing the effect of a
roll motion feedback driven steer motor on the dynamics. This model is
equivalently valid for on-road or treadmill riding~\cite{Kooijman2009a}, which
have the same fundamental dynamics. The linear version of this model can be
described by the fourth order state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states \(\vec{x}\) are the roll angle \(\phi\) and steer angle \(\delta\)
along with their time derivatives and the inputs \(\vec{u}\) are roll torque
\(T_\phi\) and steer torque \(T_\delta\). The state matrix \(\mathbf{A}\) is a
function of the equilibrium forward speed \(v\). It and the input matrix
\(\mathbf{B}\) are otherwise populated with expressions that are functions of
the geometric and inertial parameters of the nonholonomic multibody system made
up of four rigid bodies: two wheels, front frame, and rear frame.

@@

Based on findings from measuring riders without balance assist in
\cite{Haitjema2023a}, we calculate several variables that we hypothesize may
influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded perturbation force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left(F_r + F_l\right) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[\left(F_{rf} -
    F_{rr}\right) + \left(F_{lf} - F_{lr}\right)\right] dt
  \label{eq:angular-impulse}
\end{align}
%
where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
of peak torque to normalize for the duration of the applied perturbation to
capture the total effect of the perturbation pulse.

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles, \(\delta_0\) and \(\phi_0\), to
characterize the configuration of the bicycle when perturbed. The sense of
these variables relative to the sense of the perturbation matters. For example,
if there is already a large steer angle in the same direction as the
perturbation, this will have a different effect on the probability that a fall
will occur than if the steer angle is in the opposite direction of the
perturbation. To account for this, the roll and steer variables at the instance
of perturbation are normalized for perturbation direction. The gain factor
setting on the balance assist controller indicates if the assistance is off
\(\kappa=0\) or on at two different levels: low \(\kappa=3.9\) or high
\(\kappa=5.2\). As mentioned earlier, a recovery from the perturbation is
successful if the person neither places their foot down onto the treadmill
surface nor the wheel of the bicycle exits the width of the treadmill belt.  We
record this as a binary variable \(f\) for ``fall''.  All measured variables
are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain Factor & \(\kappa\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} &  DF30 wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
participant \(i\) which follows a Bernoulli distribution given the probability
\(P_{ij}\) that a fall occurs:
%
\begin{align}
  f_{ij} | P_{ij} \sim \textrm{Bernoulli}(P_{ij}) \textrm{.}
\end{align}
%
We evaluate our hypothesis using a multivariate logistic regression model that
represents the probability of this binary outcome. The log-odds of the
probability is then a linear function of our independent variables with
\(\beta\) being the intercept and \(\alpha_k\) the linear coefficients to the
\(K\) independent variables \(x_{ij}^k\), i.e. all variables in
Table~\ref{tab:stat-model-variables} except \(f\).
%
\begin{align}
  \log \left(\frac{P_{ij}}{1-P_{ij}} \right) =
  \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
  \label{eq:log-regress}
\end{align}

@@

\subsection{Utility of the Logistic Model and Extrapolation to Natural Falls}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain some
insight.
To interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall:
\(e^{\alpha_k}=0.53\).
This means if the odds are a 1000:1, turning on the balance assist system
reduces the odds to 500:1.
However, in that case the probability that a fall occurs is only reduced from
$0.999$ to $0.998$.
If the odds that a fall occurs are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse the rider was subjected to.
This means that the balance assist is most effective for perturbations close to
the participant's perturbation resistance threshold and that large
perturbations will make you fall regardless of the balance assist system's
help.

@@

This illustration alludes to the difficulty needed to apply the model in way
that could predict how many falls may be averted in a natural setting if the
balance assist system is used. Estimates of the predictor variables extracted
from the limited data collected from natural cycling would be needed to
populate the model.

The positive effect of the balance assist system is coupled to the assumptions
and experimental scenarios we implemented and there is unfortunately no simple
way to extrapolate our results to reductions of single-actor crashes we may see
if such a system were deployed widely to bicyclists. Although, our results do
indicate that we would see such a reduction, even if only in a class of
single-actor crashes that most resemble our experimental design. If there were
more comprehensive and detailed natural data of how people fall we could make
estimates on the number of falls reduced if everyone rode a balance assist
bicycle.

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why the balance assist system did not have a
statistically significant effect at 10~\si{\kph}. As a bicycle travels at
higher speeds, the same perturbation magnitude causes larger lateral
deviations. At 10~\si{\kph} almost all falls were due to the bicycle exiting
the maximum width of the treadmill. If the same experiment was performed on an
infinitely wide plane, the riders may have recovered from more perturbations.
At 6~\si{\kph} the riders could often recover in the allotted treadmill width
due to the smaller lateral deviations. We believe our results are very much
dependent on the two modes of falling, i.e. exiting the treadmill width or
placing a foot on the belt. On the other hand, cycle paths are a similar width
as the treadmill, so riders are often limited in width when recovering from a
fall thus exiting the treadmill width may be an appropriate measure for
indicating a fall. While the treadmill simulates narrow cycle paths, real-world
paths may offer more lateral recovery space, providing riders with additional
opportunities to regain balance after a perturbation. However, testing in such
narrow conditions is still highly relevant, as system design and validation
should focus on extreme conditions like narrow paths, because even in wider
paths, obstacles such as parked cars or barriers can limit lateral space.

\subsection{Rider Skill and Experience}
%
Each individual participant had their own perturbation resistance threshold,
i.e. the perturbation magnitude defining whether they would fall 50\% of the
time.
The value of this threshold is likely correlated with the skill and experience
of the participant at balancing a bicycle.
With this study, we cannot answer whether the balance assist system helps low
skilled participants more than highly skilled participants because that is
simply a function of how often a perturbation in a natural setting exceeds an
individual's threshold.
But, we did find that the system can help regardless of a participant's
individual balance performance.
On average, the balance assist system pushes the perturbation resistance
threshold up for all participants.
The vast majority of perturbations experienced in natural bicycling are below
an individual's perturbation resistance threshold, otherwise we would observe
bicyclists falling much more but low skilled riders may experience more
perturbations that exceed their threshold.
Determining how often certain perturbation types and magnitudes are experienced
in natural bicycling is an open research need and would help understand whether
a balance assist intervention would have a strong effect when used in the wild.

\subsection{Learning Effect}
%
An increase in the number of perturbations that a participant already
experienced, decreases the probability that a fall occurs. This is likely due
to the participants learning how to better recover from the perturbation over
the duration of the experiment. This effect is significant at both 6 and
10~\si{\kph}, and in the same order of magnitude. This suggests that the
learning effect that occurs during the experiment is not strongly dependent on
the speed but that similar experiments should consider the learning effect.

@@

\section{Conclusion}
%
Automatically controlling a steering motor in a bicycle using roll rate
feedback lowers the speed at which the bicycle is stable. This makes the
bicycle's low speed dynamics more akin to its uncontrolled high speed dynamics,
which is easier for a rider to control in balance. Perturbation forces applied
to the handlebar can cause a rider to fall and every rider has a perturbation
resistance threshold at which they are more likely to fall than not. The
probability of falling when mechanically perturbed is significantly
(\(p=0.02\)) reduced when traveling at 6~\si{\kph} when the balance assisting
control is activated. This effect is present when traveling at 10~\si{\kph} and
is very close to significance (\(p=0.07\)) but more investigation is needed to
determine if the effect can be significant. The positive effect to balance is
rider independent and most effective in the regime of perturbations near the
limits of the rider's control authority. Given that similar effects cause falls
during bicycling, use of the balance assist system in real world use cases will
reduce the number of falls at low bicycling speeds.

>>>>

\section*{Highlights}
%
\begin{itemize}
  \item A bicycle steering motor with automatic roll rate feedback control can
    stabilize the bicycle-rider system at speeds as low as 8~\si{\kph}.
  \item Every bicycle rider has their own perturbation resistance threshold
    when subjected to mechanical handlebar perturbations.
  \item Automatic balance assistance reduces the probability of falling while
    riding at 6~\si{\kph} when mechanically perturbed.
  \item Balance assist technology could reduce natural single-actor falls
    around the limits of the rider's control authority.
\end{itemize}

@@

Many more automatically balanced single-track vehicles have been demonstrated
over the last 40 years, but none have demonstrated that increasing low speed
stability can assist in human balancing and what effect it may have towards
single-actor falls. Most of these robotic bicycle and motorcycle designers did
not intend for a human rider to also control the stabilized vehicle.
Nevertheless, an automatically stabilized bicycle can be controlled by a human
rider if the motor controlled steer torque and the rider applied steer torque
act in parallel. This automatic control provides the ability to effectively
change the human-controlled plant dynamics, up to some limits. In our prior
study~\cite{Alizadehsaravi2023}, we demonstrated reduced motion during
distractions and light perturbations due to the balance assist system but
Hanakam, Wehner, and Wrede~\cite{Hanakam2023} recently showed rider
dissatisfaction with the stabilization of a similar vehicle. So, the overall
possible benefits are not yet definitively established.

The design of our balance assist system relies on the linear Carvallo-Whipple
bicycle model~\cite{Carvallo1899,Whipple1899} which is the simplest bicycle
model that exhibits both the non-minimum phase behavior ``countersteering'' and
self-stability. The bicycle model is well suited for showing the effect of a
roll motion feedback driven steer motor on the dynamics. This model is
equivalently valid for on-road or treadmill riding~\cite{Kooijman2009a}, which
have the same fundamental dynamics. The linear version of this model can be
described by the fourth order state space equations:
%
\begin{align}
  \dot{\vec{x}} = \mathbf{A} \vec{x} +
  \mathbf{B}
  \vec{u}
  \textrm{ where }
  \vec{x} = \begin{bmatrix} \phi \\ \dot{\phi} \\ \delta \\ \dot{\delta} \end{bmatrix}
  \textrm{ and }
  \vec{u} = \begin{bmatrix} T_{\phi} \\ T_{\delta} \end{bmatrix}
  \textrm{.}
\end{align}
%
The states \(\vec{x}\) are the roll angle \(\phi\) and steer angle \(\delta\)
along with their time derivatives and the inputs \(\vec{u}\) are roll torque
\(T_\phi\) and steer torque \(T_\delta\) as defined in ~\cite{Meijaard2007}.
The state matrix \(\mathbf{A}\) is a function of the equilibrium forward speed
\(v\). It and the input matrix \(\mathbf{B}\) are otherwise populated with
expressions that are functions of the geometric and inertial parameters of the
nonholonomic multibody system made up of four rigid bodies: two wheels, front
frame, and rear frame.

@@

Based on our prior findings from measuring riders without balance
assist~\cite{Haitjema2023a}, we calculate several variables that we hypothesize
may influence fall probability. We use the angular impulse \(L\) of the
perturbation forces over a 0.3~\si{\second} duration to characterize the
magnitude of delivered perturbation. The duration is selected based on the
duration of the commanded perturbation force and is calculated as follows:
%
\begin{align}
  L =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left(F_r + F_l\right) dt
  =
  \int_{0\si{\second}}^{0.3\si{\second}} \frac{l}{2}\left[\left(F_{rf} -
    F_{rr}\right) + \left(F_{lf} - F_{lr}\right)\right] dt
  \label{eq:angular-impulse}
\end{align}
%
where \(F_r\) and \(F_l\)  is the total force applied on the right and left
handlebar ends, respectively which are the sum of the rear and front load cell
readings \(F_{rr}\) and \(F_{rf}\), for example. The handlebar length is given
as \(l\) in Equation \ref{eq:angular-impulse}. We use angular impulse instead
of peak torque to normalize for the duration of the applied perturbation to
capture the total effect of the perturbation pulse.

We record the order number of each perturbation \(j\), i.e. first, second,
third, \ldots, to measure how many perturbations a rider is subjected to before
the current perturbation. At the initiation of each perturbation we log the
instantaneous steer and roll angles, \(\delta_0\) and \(\phi_0\), to
characterize the configuration of the bicycle when perturbed. The sense of
these variables relative to the sense of the perturbation matters. For example,
if there is already a large steer angle in the same direction as the
perturbation, this will have a different effect on the probability that a fall
will occur than if the steer angle is in the opposite direction of the
perturbation. To account for this, the roll and steer variables at the instance
of perturbation are normalized for perturbation direction as was done in
\cite{Haitjema2023a}. The gain factor setting on the balance assist controller
indicates if the assistance is off \(\kappa=0\) or on at two different levels:
low \(\kappa=3.9\) or high \(\kappa=5.2\). As mentioned earlier, a recovery
from the perturbation is successful if the person neither places their foot
down onto the treadmill surface nor the wheel of the bicycle exits the width of
the treadmill belt.  We record this as a binary variable \(f\) for ``fall''.
All measured variables are reported in Table~\ref{tab:raw}.
%
\begin{table}
  \centering
  \caption{Raw measurements taken during each trial. ``Fall Outcome'' and
    ``Perturbation Order Number'' are recorded per perturbation instance. The
    remaining measures are time varying during each trial.}
  \begin{tabular}{llll}
    \toprule
    Measure & Variable & Units & Sensor \\
    \midrule
    Balance Assist Gain Factor & \(\kappa\) & \si{\newton\second\squared} & NA \\
    Bicycle Speed & \(v\) & \si{\meter\per\second} &  DF30 wheel encoder \\
    Fall Outcome & \(f\) & boolean & NA \\
    Force \(l\)eft/\(r\)ight,\(f\)ront/\(r\)ear &
    \(F_{lf},F_{rf},F_{lr},F_{rr}\) & \si{\newton} & inline load cell\\
    Perturbation Order Number & \(j\) & integer & NA\\
    Roll Angle & \(\phi\) & \si{\degree} & BN0086 Kalman estimate \\
    Roll Angular Rate & \(\dot{\phi}\) & \si{\degree\per\second} & BN0086 rate gyroscope \\
    Steer Angle & \(\delta\) & \si{\degree} & steer tube encoder \\
    \bottomrule
  \end{tabular}
  \label{tab:raw}
\end{table}

@@

Fall outcome \(f_{ij}\) is the binary outcome of perturbation \(j\) on
participant \(i\) which follows a Bernoulli distribution given the probability
\(P_{ij}\) that a fall occurs:
%
\begin{align}
  f_{ij} | P_{ij} \sim \textrm{Bernoulli}(P_{ij}) \textrm{.}
\end{align}
%
We evaluate our hypothesis using a multivariate logistic regression model,
Equation~\ref{eq:log-regress}, that represents the probability of this binary
outcome. The log-odds of the probability is then a linear function of our
independent variables with \(\beta\) being the intercept and \(\alpha_k\) the
linear coefficients to the \(K\) independent variables \(x_{ij}^k\), i.e. all
variables in Table~\ref{tab:stat-model-variables} except \(f\).
%
\begin{align}
  \log \left(\frac{P_{ij}}{1-P_{ij}} \right) =
  \beta + \sum_{k=0}^{K} \alpha_k x_{ij}^{k}
  \label{eq:log-regress}
\end{align}

@@

\subsection{Utility of the Logistic Model to Natural Falls}
%
The probability that a fall occurs depends on the values of all the independent
variables in Table~\ref{tab:stat-model-variables}, but we can visualize the
effect of one or two variables (e.g. Figure~\ref{fig:probability}) to gain some
insight.
To interpret the results in Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10} it is important to understand the relationship between
probability and odds.
The estimate in Table~\ref{tab:freq-coefs-6} shows that the balance assist
system halves the odds that a perturbation results in a fall:
\(e^{\alpha_k}=0.53\).
This means if the odds are a 1000:1, turning on the balance assist system
reduces the odds to 500:1.
However, in that case the probability that a fall occurs is only reduced from
$0.999$ to $0.998$.
If the odds that a fall occurs are smaller, halving the odds has a larger
influence on the fall probability.
For example, if the odds that a fall occurs is two, halving it to one reduces
the probability from $0.66$ to $0.50$.
This can be seen in Figure~\ref{fig:probability} which shows the fall
probability as a function of the magnitude of the normalized angular impulse
for both speeds.
The difference in probability between the balance assist on/off case becomes
insignificant outside of approximately \(\pm1\) standard deviation of the
average angular impulse the rider was subjected to.
This means that the balance assist is most effective for perturbations close to
the participant's perturbation resistance threshold and that large
perturbations will make you fall regardless of the balance assist system's
help.

@@

This illustration alludes to the difficulty needed to apply the model in way
that could predict how many falls may be averted in a natural setting if the
balance assist system is used. Estimates of the predictor variables extracted
from the limited data collected from natural cycling would be needed to
populate the model. The positive effect of the balance assist system is coupled
to the assumptions and experimental scenarios we implemented and there is
unfortunately no simple way to extrapolate our results to reductions of
single-actor crashes we may see if such a system were deployed widely to
bicyclists. Although, our results do indicate that we would see such a
reduction, even if only in a class of single-actor crashes that most resemble
our experimental design. If there were more comprehensive and detailed natural
data of how people fall we could make estimates on the number of falls reduced
if everyone rode a balance assist bicycle.

\subsection{Treadmill Width}
%
Angular impulse magnitude has the largest significant effect for predicting
fall probability as seen in both Tables~\ref{tab:freq-coefs-6} and
\ref{tab:freq-coefs-10}. An increase in angular impulse increases the fall
probability both at 6 and 10~\si{\kph}. At 10~\si{\kph}, the multiplicative
change in odds is approximately twice as big as at 6~\si{\kph}. Thus, angular
impulse is a more important predictor at higher speeds compared to lower
speeds. We posit that this likely has to do with the width of the treadmill and
that this could also be why the balance assist system did not have a
statistically significant effect at 10~\si{\kph}. As a bicycle travels at
higher speeds, the same perturbation magnitude causes larger lateral
deviations. At 10~\si{\kph}, almost all falls were due to the bicycle exiting
the maximum width of the treadmill. If the same experiment was performed on an
infinitely wide plane, the riders may have recovered from more perturbations.
At 6~\si{\kph}, the riders could often recover in the allotted treadmill width
due to the smaller lateral deviations. We believe our results are very much
dependent on the two modes of falling, i.e. exiting the treadmill width or
placing a foot on the belt. Future studies should consider coding fall mode by
type. On the other hand, cycle paths are a similar width as the treadmill, so
riders are often limited in width when recovering from a fall thus exiting the
treadmill width may be an appropriate measure for indicating a fall. While the
treadmill simulates narrow cycle paths, real-world paths may offer more lateral
recovery space, providing riders with additional opportunities to regain
balance after a perturbation. However, testing in such narrow conditions is
still highly relevant, as system design and validation should focus on extreme
conditions like narrow paths, because even in wider paths, obstacles such as
parked cars or barriers can limit lateral space.

\subsection{Rider Skill and Experience}
%
Each individual participant had their own perturbation resistance threshold,
i.e. the perturbation magnitude defining whether they would fall 50\% of the
time.
The value of this threshold is likely correlated with the skill and experience
of the participant at balancing a bicycle.
With this study, we cannot answer whether the balance assist system helps low
skilled participants more than highly skilled participants because that is
simply a function of how often a perturbation in a natural setting exceeds an
individual's threshold.
But, we did find that the system can help regardless of a participant's
individual balance performance.
On average, the balance assist system pushes the perturbation resistance
threshold up for all participants.
The vast majority of perturbations experienced in natural bicycling are below
an individual's perturbation resistance threshold, otherwise we would observe
bicyclists falling much more, but low skilled riders may experience more
perturbations that exceed their threshold.
Determining how often certain perturbation types and magnitudes are experienced
in natural bicycling is an open research need and would help understand whether
a balance assist intervention would have a strong effect when used in the wild.

\subsection{Learning Effect}
%
An increase in the number of perturbations that a participant already
experienced, decreases the probability that a fall occurs. This is likely due
to the participants learning how to better recover from the perturbation over
the duration of the experiment. This effect is significant at both 6 and
10~\si{\kph}, and in the same order of magnitude. This suggests that the
learning effect that occurs during the experiment is not strongly dependent on
the speed but that similar experiments should consider the learning effect. An
experiment that tests professional cyclists or one that trains each participant
to their maximum performance could be insightful.

@@

\section{Conclusion}
%
Automatically controlling a steering motor in a bicycle using roll rate
feedback lowers the speed at which the bicycle is stable. This makes the
bicycle's low speed dynamics more akin to its uncontrolled high speed dynamics,
which is easier for a rider to control in balance. Perturbation forces applied
to the handlebar can cause a rider to fall and every rider has a perturbation
resistance threshold at which they are more likely to fall than not. The
probability of falling when mechanically perturbed is significantly
(\(p=0.02\)) reduced when traveling at 6~\si{\kph} when the balance assisting
control is activated. This effect is present when traveling at 10~\si{\kph} and
is very close to significance (\(p=0.07\)) but more investigation is needed to
determine if the effect can be significant. The positive effect to balance is
rider independent and most effective in the regime of perturbations near the
limits of the rider's control authority. Given that similar effects cause falls
during bicycling, use of the balance assist system in real world use cases
should reduce the number of falls at low bicycling speeds.
